---
title: "Lasso and Elastic-Net Regularized for Logistic Regression"
author: "Chunyi Wu and Andrzej Galecki"
date: "`r as.character(Sys.Date(), format = '%A %B %d, %Y')`"
output:
  rmdformats::readthedown:
    lightbox: true
    use_bookdown: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment="#>")
```

```{r, data, echo=FALSE, message=FALSE, warning=FALSE}
if (!require('pacman')) install.packages('pacman', repos = "http://cran.us.r-project.org")
library(pacman)

pacman::p_load(
  readxl,       # load excel file
  glmnet,       # Lasso and Elastic-Net Regularized Generalized Linear Models
  rmdformats,   # rmd formats
  rmarkdown,    # rmarkdown
  here,         # File locator
  skimr,        # get overview of data
  tidyverse,    # data management + ggplot2 graphics 
  tidymodels,   # for the recipes package, along with the rest of tidymodels
  janitor,      # adding totals and percents to tables
  flextable,    # format the output table
  gtsummary,    # calculate summary statistics and format the results
  sjPlot,       # correlation matrix
  purrr,        # enhances R functional programming (FP) toolki 
  tidyr,        #Tools to help to create tidy data
  ggplot2,      #Plot the results
  glmnetUtils,  #Glmnet models for multiple alpha
  coefplot      # Plotting Model Coefficients
  )
```
https://www.tidymodels.org/start/recipes/#data-split
# Introduction

In this report we consider two logistic regression models for the  `status` (0/1) variable:

* M1: Contains 21 proteins as candidate covariates
* M2 :Contains 21 prteins and log10(Baseline HbA1c), log10(ACR), eGFR as candidate covariates

We will refer to Model M1 as "unadjusted" and to Model M2 as "adjusted" model.




# Read Data

Data are prepared for "unadjusted" logistic regression (Model M1)

```{r, readxl-data, echo=TRUE, message=FALSE, warning=FALSE}
#read the data that is stored under the data folder

xlsx_path = "./datain/data_example.xlsx"
data_orig <- readxl::read_excel(xlsx_path, na = "", guess_max = 1000)

dim(data_orig)
colnames(data_orig)
# Vectors with variable names in data_orig 
protein_nms <- c(
 "KIM1.npx",       "SYND1.npx",   "IL.1RT1.npx", "WFDC2.npx", "CD27.npx", 
 "TNFRSF10A.npx",  "LAYN.npx", "PVRL4.npx",   "EDA2R.npx", "TNFRSF4.npx", 
 "GFR_alpha_1_npx","TNF.R1.npx",  "PI3_npx",  "EFNA4.npx", "TNF.R2.npx",
 "DLL1.npx",       "TNFRSF6B.npx","CD160.npx",   "EPHA2.npx", "RELT.npx",
   "LTBR.npx")
varx1 <- c("B_HBA1C_PRC", "DU_ACR", "BL_eGFR")
tvars <- c("CASE_CONTROL","FU_TIME")
varx2  <- c("SEX", "SBP_TL", "DBP_TL", "AGEONSET_TL", "BMI_TL", "AGE_TL")
vars_sel <- c("INDEX",tvars, protein_nms, varx1, varx2)

data_all <- data_orig  %>%  
  select(all_of(vars_sel)) %>%
  rename(., status = CASE_CONTROL, time = FU_TIME) %>% # rename variables
  mutate(status = factor(status))  
```

```{r data-all-summ}
dim(data_all)
colnames(data_all)   
glimpse(data_all)
```
```{r load cva-objects}
load("./datain/21Logistic-Reg1-saveFin1.Rdata") # `cva_pfit1`
load("./datain/21Logistic-Reg1-saveFin2.Rdata") # `cva_pfit2`
```

... To be continued
