---
title: 'Regularized Cox Regression Model Performance'
author: "Chunyi Wu and Andrzej Galecki"
date: "`r as.character(Sys.Date(), format = '%A %B %d, %Y')`"
output:
  rmdformats::readthedown:
    lightbox: true
    use_bookdown: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment="#>")
```

```{r, data, echo=FALSE, message=FALSE, warning=FALSE}
if (!require('pacman')) install.packages('pacman', repos = "http://cran.us.r-project.org")
library(pacman)

pacman::p_load(
  readxl,       # load excel file
  glmnet,       # Lasso and Elastic-Net Regularized Generalized Linear Models
  rmdformats,   # rmd formats
  rmarkdown,    # rmarkdown
  here,         # File locator
  skimr,        # get overview of data
  tidyverse,    # data management + ggplot2 graphics 
  janitor,      # adding totals and percents to tables
  flextable,    # format the output table
  gtsummary,    # calculate summary statistics and format the results
  sjPlot,       # correlation matrix
  purrr,        # enhances R functional programming (FP) toolki 
  tidyr,        #Tools to help to create tidy data
  ggplot2,      #Plot the results
  glmnetUtils,  #Glmnet models for multiple alpha
  coefplot,     # Plotting Model Coefficients
  survival,     #survival model 
  tidymodels,   #for modeling and machine learning using tidyverse principles
  survivalROC   #survivalROC
  )
```

# Introduction

In this report we consider two Cox regression models for the  time-to-event `time` variable. Originally this 
variable was named `FU_TIME`.  

`status` (0/1) variable, originally named `CASE_CONTROL`, is coded 0 for Controls (i.e., ESKD event has not occurred), and 1 for Cases (i.e., event has occurred)

* M1: Contains 21 proteins as candidate covariates
* M2 :Contains 21 proteins and Baseline HbA1c, log10(ACR), BL_eGFR, SEX, and AGE_TL(Baseline Age) as candidate covariates

We will refer to Model M1 as "unadjusted" and to Model M2 as "adjusted" model.

# Auxiliary functions

* Auxilary function `get_model_params()` defined.

```{r get-model-params-fun}
# Extract the best (hyper) parameters from `cva.glmnet` object.
get_model_params <- function(cvafit) {
  alpha <- cvafit$alpha
  lambdaMin <- sapply(cvafit$modlist, `[[`, "lambda.min")
  lambdaSE <- sapply(cvafit$modlist, `[[`, "lambda.1se")
  error <- sapply(cvafit$modlist, function(mod) {min(mod$cvm)})
  best <- which.min(error)
  data.frame(alpha = alpha[best], lambdaMin = lambdaMin[best],
             lambdaSE = lambdaSE[best], eror = error[best])
}
```

* `survivalROC_helper()` function is defined:
* source: (https://datascienceplus.com/time-dependent-roc-for-survival-prediction-models-in-r)


```{r survivalROC-helper-function}
## Define a helper function to evaluate at various t
survivalROC_helper <- function(t, markr) {
    survivalROC(Stime        = data_test$time,
                status       = data_test$status,
                marker       = data_test[[markr]],
                predict.time = t,
                method       = "NNE",
                span = 0.25 * nrow(data_test)^(-0.20))
}
```


* Auxiliary `plot_timedep_ROC()` function to Plot Time-dependent ROC is defined

```{r surv-roc-plot-function}

## Plot Time-dependent ROC function

plot_timedep_ROC <- function(surv_ROC_data){
 surv_ROC_data %>%
    ggplot(mapping = aes(x = FP, y = TP)) +
    geom_point() +
    geom_line() +
    geom_label(data = surv_ROC_data %>% select(t,auc) %>% unique,
               mapping = aes(label = sprintf("%.3f", auc)), x = 0.5, y = 0.5) +
    facet_wrap( ~ t) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
    }     
```

# Data preparation 

## Read Raw Data  

Original data are stored in Excel format.
Data are used for external validation.

are prepared for "unadjusted" Cox regression (Model M1). 

```{r, read-data, echo = TRUE, message=FALSE, warning=FALSE}

xlsx_path <- "./datain/data_example.xlsx"

#read the data that is stored under the data folder

data_original <- readxl::read_excel(xlsx_path, guess_max = 1000)
dim(data_original) # No. of rows and columns


data0 <- data_original  %>% mutate(status=CASE_CONTROL, time=FU_TIME) %>%
    mutate(log10_DU_ACR=log10(DU_ACR))  %>% filter(time>0)

dim(data0) # Number of rows and columns in `data0`
glimpse(data0)
```

Remove rows with missing values from intermediate dataset

```{r data-drop-NA}
prot_npx <- c("KIM1.npx",     "SYND1.npx", "IL.1RT1.npx",   "WFDC2.npx", "CD27.npx",
              "TNFRSF10A.npx","LAYN.npx","PVRL4.npx",   "EDA2R.npx","TNFRSF4.npx",
              "GFR_alpha_1_npx","TNF.R1.npx","PI3_npx", "EFNA4.npx","TNF.R2.npx" ,
              "DLL1.npx", "TNFRSF6B.npx", "CD160.npx", "EPHA2.npx","RELT.npx",
              "LTBR.npx") 
surv_vars <-  c("time", "status")
clin_xvars <- c("B_HBA1C_PRC", "log10_DU_ACR", "BL_eGFR", "SEX", "AGE_TL") # Clinical vars used in M2
xvars1 <- c(surv_vars, prot_npx, clin_xvars)
 
data <- data0  %>% select(all_of(xvars1)) 

#drop any records with NA
data_drop_na <- data %>% drop_na()    

# Total sample size and number of columns
dim(data_drop_na)
```

* time and status in `data_drop_na`

```{r time-status-NA}
# time and status
tx <- data.matrix(data_drop_na[,c("time","status")])
ty    <- tx[, "time"]
stx <- tx[,"status"] 
ySurv <- survival::Surv(ty, stx)   
table(data_drop_na$status) # 0-censored, 1-observed time-to-ESKD
```

## Data for testing



```{r prepare-test-data}
test_rows <- 1:length(ty)    # Select rows for testing. Possibly all rows in data_drop_na.
y_test <- ty[test_rows]
stx_test <-stx[test_rows]
ySurv_test <- survival::Surv(y_test, stx_test)
summary(ySurv_test)
data_test <- data_drop_na[test_rows, ]
```

# Model M1 

Prepare model M1 matrices and extract

## Model M1 newx matrix

Model M1 design matrix `x1_test` is prepared for external validation/testing.


```{r model-M1-xmtx}
#includes all proteins after excluding missing data
x1_test <- data.matrix(data_test[, prot_npx])
dim(x1_test)
colnames(x1_test)  # 21 proteins
```

## Load cva object for  M1

* Load cva object for  M1 fitted to training data. 
* Extract optimal combination of alpha and lambda hyperparameters

```{r cva-pfit1Fin-load}
load("./datain/21Cox-Reg1-saveFin1.Rdata")
minlossplot(cva_pfit1)
```



Extract the optimal (hyper)parameters from `cva.glmnet` object

```{r get_best_params}
get_model_params(cva_pfit1)
a1_opt <- get_model_params(cva_pfit1)[["alpha"]]
lmbda1_opt <- get_model_params(cva_pfit1)[["lambdaMin"]]
```

Cox regression coefficients for Model M1 using the optimal combination of (hyper)parameters

```{r get_best_coef}
coef(cva_pfit1, s = lmbda1_opt, alpha = a1_opt)
```

## Extract optimal M1 fit

* Extract optimal M1 fit from cva object

```{r pfit1_aopt-object}

which_alpha    <- which(cva_pfit1$alpha == a1_opt)  # Position of the optimal alpha
cv_pfit1_aopt  <- cva_pfit1$modlist[[which_alpha]]
pfit1_aopt     <- cv_pfit1_aopt$glmnet.fit  # glmnet fit(s) for the optimal alpha
class(pfit1_aopt)
```

## Predictive performance of optimal M1

###  predicted values for test data


* Calculate predicted values for test data

```{r calc-pred1-vals}
predM1_lpmtx <- predict(pfit1_aopt, newx = x1_test, type = "link") # Matrix of predicted values (multiple lambdas)
predM1_lp <- predict(pfit1_aopt, newx = x1_test, type = "link", s = lmbda1_opt)
```


### C-index

Ref: Harrel Jr, F. E. and Lee, K. L. and Mark, D. B. (1996) Tutorial in biostatistics:
multivariable prognostic models: issues in developing models, evaluating assumptions 
and adequacy, and measuring and reducing error, _Statistics in Medicine_, 15, pages 361-387.


```{r C-index-M1}
apply(predM1_lpmtx, 2, Cindex, y = ySurv_test) # Multiple lambdas
Cindex(predM1_lp, ySurv_test)                  # Optimal lambda
```

### Time-dependent ROC for M1



```{r surv-roc-M1}
                
## Augment `data_test` with linear predictor for M1
 data_test$lp_M1 <- predM1_lp
 

## Evaluate every 2.5 years
survivalROC_data1 <- tibble(t = c(5,7.5, 10, 12.5,15)) %>%
    mutate(survivalROC = map(t, survivalROC_helper, markr = "lp_M1"),
           ## Extract scalar AUC
           auc = map_dbl(survivalROC, magrittr::extract2, "AUC"),
           ## Put cut off dependent values in a data_frame
           df_survivalROC = map(survivalROC, function(obj) {
               as_tibble(obj[c("cut.values","TP","FP")])
           })) %>%
           select(-survivalROC) %>%
    unnest(df_survivalROC) %>%
    arrange(t, FP, TP)
    
survivalROC_data1 %>% print (n=10)
```

Plot Time-dependent ROC for M1

```{r surv-roc-plot-M1}
plot_timedep_ROC(survivalROC_data1)
```

# Model M2 

Prepare model M2 matrices and extract M2 from cva object

## Model M2 newx matrix

Model M2 design matrix `x2_test` is prepared for external validation/testing.


```{r model-M2-xmtx}
#includes all proteins after excluding missing data
x2_test <- data.matrix(data_test[, c(prot_npx, clin_xvars)])
dim(x2_test)
colnames(x2_test)  # 21 proteins
```

## Load cva object for  M2

^ Load cva object for  M2 fitted to training data. 
* Extract optimal combination of alpha and lambda hyperparameters

```{r cva-pfit2Fin-load}
load("./datain/21Cox-Reg1-saveFin2.Rdata")
minlossplot(cva_pfit2)
```

* Auxilary function `get_model_parms()` defined earlier


Extract the optimal (hyper)parameters from `cva.glmnet` object

```{r get_best_params2}
get_model_params(cva_pfit2)
a2_opt <- get_model_params(cva_pfit2)[["alpha"]]
lmbda2_opt <- get_model_params(cva_pfit2)[["lambdaMin"]]
```

Cox regression coefficients for Model M2 using the optimal combination of (hyper)parameters

```{r get_best_coef2}
coef(cva_pfit2, s = lmbda2_opt, alpha = a2_opt)
```

## Extract optimal M2 fit

* Extract optimal M2 fit from cva object

```{r pfit2_aopt-object}

which_alpha    <- which(cva_pfit2$alpha == a2_opt)  # Position of the optimal alpha
cv_pfit2_aopt  <- cva_pfit2$modlist[[which_alpha]]
pfit2_aopt     <- cv_pfit2_aopt$glmnet.fit  # glmnet fit(s) for the optimal alpha
class(pfit2_aopt)
```

## Predictive performance of optimal M2

###  predicted values for test data


* Calculate predicted values for test data

```{r calc-pred2-vals}
predM2_lpmtx <- predict(pfit2_aopt, newx = x2_test, type = "link") # Matrix of predicted values (multiple lambdas)
predM2_lp    <- predict(pfit2_aopt, newx = x2_test, type = "link", s = lmbda2_opt)
```


### C-index

Ref: Harrel Jr, F. E. and Lee, K. L. and Mark, D. B. (1996) Tutorial in biostatistics:
multivariable prognostic models: issues in developing models, evaluating assumptions 
and adequacy, and measuring and reducing error, _Statistics in Medicine_, 15, pages 361-387.


```{r C-index-M2}
apply(predM2_lpmtx, 2, Cindex, y = ySurv_test) # Multiple lambda
Cindex(predM2_lp, ySurv_test)                  # Optimal lambda
```

### Time-dependent ROC for M2

* source: (https://datascienceplus.com/time-dependent-roc-for-survival-prediction-models-in-r)



```{r surv-roc-M2}
                
## Augment `data_test` with linear predictor for M2
 data_test$lp_M2 <- predM2_lp
 
## Helper function `survivalROC_helper()` defined earlier

## Evaluate every 2.5 years
survivalROC_data <- tibble(t = c(5, 7.5, 10, 12.5,15)) %>%
    mutate(survivalROC = map(t, survivalROC_helper, markr = "lp_M2"),
           ## Extract scalar AUC
           auc = map_dbl(survivalROC, magrittr::extract2, "AUC"),
           ## Put cut off dependent values in a data_frame
           df_survivalROC = map(survivalROC, function(obj) {
               as_tibble(obj[c("cut.values","TP","FP")])
           })) %>%
           select(-survivalROC) %>%
    unnest(df_survivalROC) %>%
    arrange(t, FP, TP)
    
survivalROC_data %>% print (n=10)
```

* Plot Time-dependent ROC every 3 years

```{r surv-roc-plot}
## Plot Time-dependent ROC every 3 years
survivalROC_data %>%
    ggplot(mapping = aes(x = FP, y = TP)) +
    geom_point() +
    geom_line() +
    geom_label(data = survivalROC_data %>% select(t,auc) %>% unique,
               mapping = aes(label = sprintf("%.3f", auc)), x = 0.5, y = 0.5) +
    facet_wrap( ~ t) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
          
```






